Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2015-06-04 tex/camera_ready/paper.tex
shortened

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples to formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{longv}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}%
\begin{shortv}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{shortv}%
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
    (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
      (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
        (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
          (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
            (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14))
\end{verbatim}
\end{quote}

@@

\begin{longv}
Most of the above tools cannot cope with algebraic datatypes or other infinite
types. \end{longv}%
Kuncak and Jackson \cite{kuncak-jackson-2005} presented an idiom for
encoding datatypes and recursive functions in Alloy, by approximating datatypes
by finite subterm-closed substructures. The approach finds sound (fragments
of) models for formulas in the \relax{existential--bounded-universal} fragment
\begin{longv}(i.e., formulas whose prenex normal forms contain no unbounded universal
quantifiers ranging over datatypes)\end{longv}. This idiom was further developed by Dunets
et al.\ \cite{dunets-et-al-2010}, who presented a translation scheme
for primitive recursion. Their definedness guards play a similar role to the
existential constraints generated by our translation $\conv$.

@@

Model finding is concerned with satisfying arbitrary logical constraints. Some
tools are tailored for problems that correspond to total functional
programs. The QuickCheck tool for Haskell \cite{claessen-hughes-2000} is an
early example, based on random testing. Bounded exhaustive testing\begin{longv}
\cite{runciman-et-al-2008}\end{longv} and narrowing\begin{longv}
\cite{lindblad-2008-testing}\end{longv} are other successful strategies. These
tools are often much faster than model finders, but they typically cannot cope
with underspecification and nonexecutable functions.

>>>>

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples to formulas involving recursive
function definitions.
For convenience, it
has been implemented in the development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the upcoming SMT-LIB 2.5 standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{longv}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{longv}%
\begin{shortv}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{shortv}%
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs a partial model of~$\con{s}$, where only
the values of $\con{s}$ for $x = 0,\ldots,14$ are given.

@@

\begin{longv}
Most of the above tools cannot cope with algebraic datatypes or other infinite
types. \end{longv}%
Kuncak and Jackson \cite{kuncak-jackson-2005} presented an idiom for
encoding datatypes and recursive functions in Alloy, by approximating datatypes
by finite subterm-closed substructures.
%The approach finds sound (fragments
%of) models for formulas in the \relax{existential--bounded-universal} fragment
%\begin{longv}(i.e., formulas whose prenex normal forms contain no unbounded universal
%quantifiers ranging over datatypes)\end{longv}.
This idiom was extended by Dunets
et al.\ \cite{dunets-et-al-2010}, who presented a translation scheme
for primitive recursion. Their definedness guards play a similar role to our
existential constraints.

@@

Model finding is concerned with satisfying arbitrary logical constraints. Some
tools are tailored for problems that correspond to total functional
programs. The QuickCheck tool \cite{claessen-hughes-2000} for Haskell is an
early example, based on random testing. Bounded exhaustive testing\begin{longv}
\cite{runciman-et-al-2008}\end{longv} and narrowing\begin{longv}
\cite{lindblad-2008-testing}\end{longv} are other successful strategies. These
tools are often much faster than model finders, but they typically cannot cope
with underspecification and nonexecutable functions.
